TacticToe: Learning the Game of Tactics

### 2018-02-19 tactictoe_jv.tex
intergrating Lasse's comments

\paragraph{Tactics in \holfour.}
Considering concrete implementation details now, a \emph{tactic} is a function
that takes a goal (or proof obligation) and returns a list of goals (subgoals of the original goal) together
with a validation function.
Calling the validation function on a list of \emph{theorems}, corresponding to the list of subgoals, results in a theorem corresponding to the original goal.
For example, if the list of subgoals is empty, then calling the validation function on the empty list yields the originally goal sequent as a theorem.
In this way, tactics implement the plumbing of backwards, goal-directed, proof construction.
\todo{do not mention validation? RK: it depends whether these details are helpful for understanding or not}

@@

\begin{itemize}
\item Proof recording at the tactic level is made more precise. Pattern
matching constructions, opening of submodules are supported. The different
recording steps (parsing, unfolding and replaying) are regrouped into a single
\sml function.
\item Monte Carlo tree search~~\cite{montecarlo} (MCTS) replaces A* as our
search algorithm. The MCTS algorithm gives a more balanced feedback on
the success of
each proof step by
comparing subtrees of the search tree instead of leafs. The policy and
evaluation are learned
through supervised learning.
\item Proof guidance required by MCTS is given by predictors for
three kind of objects: tactics, theorems and list of goals.
\item The orthogonalization process that eliminates redundant tactics is
formally described.
\item A tactic abstraction method is introduced. It enables us to create
more general and flexible tactics where tactic arguments are dynamically
predicted.
\item The internal ATP \metis is complemented
by asynchronous calls to \eprover to help \tactictoe during proof search.
\item Evaluation of \tactictoe success rate relative to
the length of the original proofs.
\item User interaction is enhanced by the minimization and prettification
applied on the generated proof.
\end{itemize}

\section{Preliminaries}\label{sec:prelim}

The different components of \tactictoe revolve around the concept of a
search tree. The following definitions are made relative to a static search
tree, the description of its evolution during proof search will be given in
Section~\ref{sec:proofsearch}.

\begin{definition}(search tree)\\
A search tree $\mathfrak{T}$ is a sixtuple
$(\mathbb{T},\mathbb{G},\mathbb{A},T,G,A)$
that respects the following conditions:
\begin{itemize}
\item $\mathbb{T}$ is a set of tactics, $\mathbb{G}$ is a set of goals
 and $\mathbb{A}$ is a set of nodes representing lists of goals.
\item $T$ is a total function from $\mathbb{G}$ to $\mathcal{P}(\mathbb{T})$. It takes a
goal and return the set of tactics already applied to this goal.
\item $G$ is a total function from $\mathbb{A}$ to $\mathcal{P}(\mathbb{G})$. It takes a
node and returns the list of goals of this node.
\item $A$ is a partial function from $\mathbb{G} \times \mathbb{T}$ to
$\mathbb{A}$. It takes a goal and a tactic and returns the node produced by the
application of the tactic to the goal. Its support $\mathit{Support}(A)$ is by
definition $\lbrace (g,t) \in
\mathbb{G} \times \mathbb{T}\ |\ t \in T(g)
\rbrace$.
\item If $(g,t) \in \mathit{Support}(A)$, then $t(g) = G(A(g,t))$. This is the
main condition. The result of the application of a tactic $t$
to a goal $g$ is exactly the list of goals contained in the node produced by
$t$ from $g$.
\end{itemize}

Additional constraints such as acyclicity and the fact that there is exactly
one root node are properties of this graph that are preserved during the
application of the MCTS algorithm. That is why we call it a tree, although this
is not specified in this definition.
\end{definition}

If no explicit order is given, we assume that any set is equipped with an
arbitrary order. In this way, we can access the $n^{th}$ member of a set $s$ by
writing $s_n$. In Figure~\ref{fig:choice}, $(G(a_j))_i$ is written as $g_j^i$.

\begin{figure}{}
\begin{center}
\begin{tikzpicture}[node distance=0.7cm]
\node [tcircle] (0) {$g_0^i$};
\node [right of=0] (0r) {.\ .\ .};
\node [tcircle,right of=0r] (0rr) {$g_0^n$};
\node [left of=0] (0l) {.\ .\ .};
\node [tcircle,left of=0l] (0ll) {$g_0^1$};
\node [draw,fit=(0rr) (0ll)] (0b) {};
\node [left of=0ll] {$a_0$};
\node [above of=0,node distance=2cm] (2) {.\ .\ .};
\node [left of=2, tcircle] (2l) {$g_j^1$};
\node [tcircle,right of=2] (2r) {$g_j^p$};
\node [draw, fit=(2l) (2r)] (2b) {};
\node [left of=2l] {$a_j$};

\node [tcircle, left of=2l, node distance=3cm] (1r) {$g_1^q$};
\node [left of=1r] (1) {.\ .\ .};
\node [tcircle, left of=1] (1l) {$g_1^1$};
\node [draw, fit=(1l) (1r)] (1b) {};
\node [left of=1l] {$a_1$};

\node [right of=2r,node distance=3cm] (3l) {$\phantom{g_3^p}$};
\node [right of=3l] (3) {\phantom{.\ .\ .}};
\node [right of=3] (3r) {$\phantom{g_3^p}$};
\node [fit=(3l) (3r)] (3b) {};

@@

\draw[-to,thick] (0) to node[xshift=-10](t1){$t_1$} (1b);
\draw[-to,thick] (0) to node[xshift=-10](t2){$t_j$} (2b);
\draw[-to,dotted,thick] (0) to (1m);
\draw[-to,dotted,thick] (0) to (2m);
\draw[-to,dotted,thick] (0) to (3b);
\end{tikzpicture}
\end{center}
\caption{A node of a search tree and the children of one of
its goal.}
\label{fig:choice}
\end{figure}

Starting from the parent node, we can construct children of a goal of
this node by choosing a goal $g_i^0$ on this node and choosing tactics
$t_1,t_2,\ldots$ to
apply to this goal. Here the result produces new nodes each containing a list
of goals to be proven called obligations. In each node of the proof tree,
a goal can either be
solved, open or pending (see Definition~\ref{def:solved}).

In the following, concepts and properties are defined in the context of a
search tree $\mathfrak{T}$.

\begin{definition}\label{def:solved}(Solved goal, solved nodes, open goal,
pending goal)\\
The set of solved nodes $\mathbb{A}^\infty$ and
the set of solved goals $\mathbb{G}^\infty$ are defined inductively by:

\begin{align*}
\mathbb{A}^{0} &=_{def}
\lbrace a \in \mathbb{A}\ |\ G(a) = \emptyset \rbrace \\
\mathbb{G}^{0} &=_{def} \lbrace g \in \mathbb{G}\ |\
\exists a \in A(g).\ a \in \mathbb{A}^{0} \rbrace\\
\mathbb{A}^{n+1} &=_{def} \lbrace a \in \mathbb{A}\ |\
\forall g \in G(a).\ g \in \mathbb{G}^{n} \rbrace\\
\mathbb{G}^{n+1} &=_{def} \lbrace g \in \mathbb{G}\ |\
\exists a \in A(g).\ a \in \mathbb{A}^{n+1} \rbrace \\
\mathbb{A}^\infty &=_{def} \bigcup_{n \in \mathbb{N}} \mathbb{A}^n \ \ \ \ \
\mathbb{G}^\infty =_{def} \bigcup_{n \in \mathbb{N}} \mathbb{G}^n\\
\end{align*}

An open goal is an unsolved goal that was tried at least once by a tactic. The
set of open goals can be formally defined as:
\[\lbrace g \in G\ |\  T(g) \neq \emptyset \wedge g \notin \mathbb{G}^\infty
\rbrace\]

The set of pending goals is the remaining set of goals in the tree. They are
goal that were produced by tactics that were not explored yet. This set can
also be also defined by:
\[\lbrace g \in G\ |\  T(g) = \emptyset \rbrace\]

To facilitate the description  of our algorithms we make a slight adjustment
to the definition of an open goal. If a node $a$ does not contain any open
goal, the first pending goal (according
to some preset order) of $a$ is also considered to be an open goal. In this
way, each unsolved node contains a unique open goal.
\end{definition}

\todoi{some text to add here}

\begin{definition}\label{def:desc}(Children, descendant, active node)\\
The children of a goal $g$ is a set of nodes defined by:
\[\mathit{Children}(g) = \lbrace A(g,t)\in \mathbb{A}\ |\ t \in T(g) \rbrace\]

@@

An active node is by definition a node that is not a descendant of a
solved node. Non-active nodes are not worth exploring.
\end{definition}

We can now judge the productivity of a tactic by its contribution to the
search tree.

\begin{definition} (Productive tactic)
The application of a tactic on an open goal is called productive if and only if
all the following conditions are satisfied:
\begin{itemize}
\item It does not fail or timeout. To prevent tactics from looping, we
interrupt tactics after a short amount of time (0.05 seconds).
\item It does not loop. Its output does not contain any goals that appears in
the ancestor of the node of the current
goal.
\item It is not a parallel step. Its output is not equal or a superset
of a list of goals produced from the same open goal.
\end{itemize}

@@

\section{Prediction}\label{s:prediction}
The learning-based selection of relevant lemmas significantly improves the
automation for hammers~\cite{BlanchetteGKKU16}. In \tactictoe, we use the
distance-weighted \emph{$k$ nearest-neighbour} classifier~\cite{DudaniS76}
adapted for lemma selection~\cite{ckju-pxtp13}. It allows for hundreds of
predictions per second, while maintaining very high prediction quality~\cite{femalecop}.
Here it is used
to predict three different kinds of objects:
 tactics, theorems and list of goals. The number of predicted objects $k$ is
 called the \emph{prediction radius}.

We first introduce specificities associated with the prediction of each object.
In particular, we present the dataset from which objects are selected and the
purpose of the selection process.
We then explain the construction of the similarity measure by which relevant
objects are predicted.

@@

\paragraph{Theorems as Arguments of Tactics}
We first collect a set of theorems for our theorem predictor to select from.
It includes the \holfour theorem database and theorems from the local namespace.
Predicted tactics during proof search and orthogonalization may include
tactics where arguments (list of theorems) have been erased (see Section
\ref{sec:synthesis}).
We instantiate those tactics by theorems that are syntactically the closest to
the goal as they are more likely to help.
The same algorithm selects suitable premises for ATPs integrated with
\tactictoe (see Section~\ref{sec:atp}).

\paragraph{List of Goals as Output of Tactics}
A dataset of tactic outputs is compiled during orthogonalization (see
Section~\ref{sec:evaluation}).
This set is separated into positive and negative examples.
List of goals that subsumes the result of an original tactic on a
targeted goal are considered positive. During proof search, given a list of
goals $l$ created by a tactic, we select a set of lists of goals that are most
similar to $l$. The ratio of positive examples in the selection gives us a
prior evaluation for MCTS.

\subsection{Feature Extraction}\label{sec:features}
All the three kinds of considered objects to predict correspond to (different representations of)
mathematical
formulas. Therefore, we start by describing features for terms. From there, we
extend the extraction mechanism
to goals, theorems and lists of goals. Duplicated features are always removed
so that each object has an associated set of features.

Here are the different kind of feature we extract from terms:
\begin{itemize}
\item names of constants, including the logical operators,
\item type constructors present in the types of constants and variables,
\item subterms with all variables replaced by a single place holder $V$.
\item names of the variables.
\end{itemize}

Since goals and theorems are represented by a couple of a list of terms
(assumptions) and a term (conclusion). We compute their features by extracting
the features of all terms describing them. Additionally, we distinguish between
features of the assumptions and features of the conclusion by adding a
different tag to elements of each set. Features of a list of goals are
the union of the features of each goal in the list.

\subsection{Similarity}\label{sec:predictions}
We estimate the similarity (or co-distance) between two objects $o_1$ and $o_2$
through their respective feature sets $f_1$ and $f_2$.
%The similarity function $sim$ computed by the k-NN algorithm is analogous
%to the one used in the premise selection task~\cite{ckju-pxtp13}.
The estimation is based on the frequency of the features in the intersection of
$f_1$ and $f_2$. A good rule of thumb is that the rarer the shared features
are, the more similar two goals should be. The relative rarity of each feature
can be estimated by calculating its TF-IDF weight~\cite{Jones72astatistical}.
We additionally modify the weight by raising them to the sixth power giving
even more credence
to rare features which was found experimentally optimal in \cite{ckju-pxtp13}.

The first co-distance $sim_1$ adds the weight of each shared
features to compute the total score.
In a second co-distance $sim_2$, we additionally take into account
the total number of features to reduce the seemingly unfair advantage of big
feature sets in the first scoring function.
\[sim_1 (o_1, o_2) = {\sum\nolimits_{\,f \in f_0 \cap
f_1}{\text{tfidf}(f)^{6}}}\]
\[sim_2 (o_1, o_2) = \frac{{\sum\nolimits_{\,f \in f_0 \cap
f_1}{\text{tfidf}(f)^{6}}}}
{ln (e + card(f_0) + card(f_1))}\]

In our setting, making prediction for an object $o$ consists of sorting a set
of objects by their similarity to $o$. We use $sim_1$ for tactics and theorems
and $sim_2$ for list of goals. The reason why $sim_1$ is adapted for
predicting tactics is that tactics effective on large goal $g$ are often also
suitable for a goal made of sub-formulas of $g$. The same monotonicity
heuristic can justify the use of $sim_1$ in theorems. Indeed in practice, a
large theorem is often a conjunction of theorems from the same domain.
And if a conjunction contains a formula related to a problem, then the other
conjuncts from the same domain may also help to solve it.

>>>>

\paragraph{Tactics in \holfour.}
Considering concrete implementation details now, a \emph{tactic} is a function
that takes a goal (or proof obligation) and returns a list of goals (subgoals of the original goal) together
with a validation function.
Calling the validation function on a list of \emph{theorems}, corresponding to the list of subgoals, results in a theorem corresponding to the original goal.
For example, if the list of subgoals is empty, then calling the validation function on the empty list yields the originally goal sequent as a theorem.
In this way, tactics implement the plumbing of backwards, goal-directed, proof 
construction. 
Because validation functions are only important to check the final proof, we 
omit them during the description of proof search algorithm. And we note $t(g)$ 
the lists of goals produced by the application of the tactic $t$ to the goal 
$g$.

@@

\begin{itemize}
\item Proof recording at the tactic level is made more precise. Pattern
matching constructions, opening of submodules are supported. The different
recording steps (parsing, unfolding and replaying) are regrouped into a single
\sml function.
\item Monte Carlo tree search~~\cite{montecarlo} (MCTS) replaces A* as our
search algorithm. The MCTS algorithm gives a more balanced feedback on
the success of
each proof step by
comparing subtrees of the search tree instead of leafs. The policy and
evaluation are learned
through supervised learning.
\item Proof guidance required by MCTS is given by predictors for
three kinds of objects: tactics, theorems and lists of goals.
\item The orthogonalization process that eliminates redundant tactics is
formally described.
\item A tactic abstraction method is introduced. It enables us to create
more general and flexible tactics where tactic arguments are dynamically
predicted.
\item The internal ATP \metis is complemented
by asynchronous calls to \eprover to help \tactictoe during proof search.
\item Evaluation of \tactictoe success rate relative to
the length of the original proofs.
\item User interaction is enhanced by the minimization and prettification
applied on the generated proof.
\end{itemize}

\section{Search tree}\label{sec:prelim}

The application of tactic to a goal produces a list of goals. Each of the 
created goals can in turn be the input other tactics. Moreover, it is possible 
to try multiple tactics on the same goal. In order to keep track of the 
progress made so far from the initial goal we want to solve, we organize
goals and tactics into a graph with list of goals as nodes and tactics as edges
(See Definition~\ref{def:stree}).

In this section, we only give a description of the graph at a given 
moment of the search after a certain amount of tactic applications. 
The exploration of the tree is decided by MCTS algorithm in 
Section~\ref{sec:proofsearch}, which is guided by a tactic prediction algorithm 
presented in Section~\ref{sec:features}.

\begin{definition}\label{def:stree}(search tree)\\
A search tree $\mathfrak{T}$ is a sixtuple
$(\mathbb{T},\mathbb{G},\mathbb{A},T,G,A)$
that respects the following conditions:
\begin{itemize}
\item $\mathbb{T}$ is a set of tactics, $\mathbb{G}$ is a set of goals
 and $\mathbb{A}$ is a set of nodes representing lists of goals.
\item $T$ is a function from $\mathbb{G}$ to $\mathcal{P}(\mathbb{T})$. It 
takes a goal and return the set of tactics already applied to this goal.
\item $G$ is a total function from $\mathbb{A}$ to $\mathcal{P}(\mathbb{G})$. 
It takes a node and returns the list of goals of this node.
\item $A$ is a partial function from $\mathbb{G} \times \mathbb{T}$ to
$\mathbb{A}$. It takes a goal and a tactic and returns the node produced by the
application of the tactic to the goal. Its support $\mathit{Support}(A)$ is 
defined as $\lbrace (g,t) \in \mathbb{G} \times \mathbb{T}\ |\ t \in T(g) 
\rbrace$. And if $(g,t) \in \mathit{Support}(A)$, then $t(g) = G(A(g,t))$. This 
means that the output $t(g)$ is exactly the list 
of goals contained in the node produced by
$t$ from $g$.
\end{itemize}

Acyclicity and the fact that there is exactly
one root node containing only the initial goal are additional constraints to 
the search tree that are preserved during the
application of the MCTS algorithm. That is why we call it a tree, although this
is not specified in this definition. 
\end{definition}

If no explicit order is given, we assume that the sets 
$\mathbb{T},\mathbb{G},\mathbb{A}$ are equipped with an
arbitrary order. In Figure~\ref{fig:choice}, a part of the a search tree is 
depicted, with goals represented by circles, nodes by rectangles and tactics by 
arrows.

\begin{figure}{}
\begin{center}
\begin{tikzpicture}[node distance=0.7cm]
\node [tcircle] (0) {$g_i$};
\node [right of=0] (0r) {.\ .\ .};
\node [tcircle,right of=0r] (0rr) {$g_n$};
\node [left of=0] (0l) {.\ .\ .};
\node [tcircle,left of=0l] (0ll) {$g_0$};
\node [draw,fit=(0rr) (0ll)] (0b) {};
\node [left of=0ll] {$a_0$};
\node [above of=0,node distance=2cm] (2) {.\ .\ .};
\node [left of=2, tcircle] (2l) {$\phantom{g_3}$};
\node [tcircle,right of=2] (2r) {$\phantom{g_3}$};
\node [draw, fit=(2l) (2r)] (2b) {};
\node [left of=2l] {$a_j$};

\node [tcircle, left of=2l, node distance=3cm] (1r) {$\phantom{g_3}$};
\node [left of=1r] (1) {.\ .\ .};
\node [tcircle, left of=1] (1l) {$\phantom{g_3}$};
\node [draw, fit=(1l) (1r)] (1b) {};
\node [left of=1l] {$a_1$};

\node [right of=2r,node distance=3cm] (3l) {$\phantom{g_3}$};
\node [right of=3l] (3) {\phantom{.\ .\ .}};
\node [right of=3] (3r) {$\phantom{g_3}$};
\node [fit=(3l) (3r)] (3b) {};

@@

\draw[-to,thick] (0) to node[xshift=-10](t1){$t_1$} (1b);
\draw[-to,thick] (0) to node[xshift=-4](t2){$t_j$} (2b);
\draw[-to,dotted,thick] (0) to (1m);
\draw[-to,dotted,thick] (0) to (2m);
\draw[-to,dotted,thick] (0) to node[xshift=15](t2){$t_m$} (3b);
\end{tikzpicture}
\end{center}
\caption{A node of a search tree $a_0$, the list of goals $g_0 \ldots g_n$ it 
contains and the nodes $a_1 \ldots a_m$ derived from the application of the 
tactics $t_1 \ldots t_m$ to $g_i$.}
\label{fig:choice}
\end{figure}

In the following, properties are defined in the context of a
search tree $\mathfrak{T}$. In Definition~\ref{def:solved}, we specify 
different status for a goal: open, pending or solved. There, we define 
what a solved goal and solved node are by mutual recursion on the number of
steps it takes to solve a goal and a node.
\begin{definition}\label{def:solved}(Solved goal, solved nodes, open goal,
pending goal)\\
The set of solved nodes $\mathbb{A}^\infty$ and
the set of solved goals $\mathbb{G}^\infty$ are defined inductively by:

\begin{align*}
\mathbb{A}^{0} &=_{def}
\lbrace a \in \mathbb{A}\ |\ G(a) = \emptyset \rbrace \\
\mathbb{G}^{0} &=_{def} \lbrace g \in \mathbb{G}\ |\
\exists t \in T(g).\ A(g,t) \in \mathbb{A}^{0} \rbrace\\
\mathbb{A}^{n+1} &=_{def} \lbrace a \in \mathbb{A}\ |\
\forall g \in G(a).\ g \in \mathbb{G}^{n} \rbrace\\
\mathbb{G}^{n+1} &=_{def} \lbrace g \in \mathbb{G}\ |\
\exists t \in T(g).\ A(g,t) \in \mathbb{A}^{n+1} \rbrace \\
\mathbb{A}^\infty &=_{def} \bigcup_{n \in \mathbb{N}} \mathbb{A}^n \ \ \ \ \
\mathbb{G}^\infty =_{def} \bigcup_{n \in \mathbb{N}} \mathbb{G}^n\\
\end{align*}

We call a goal (respectively node) unsolved if it is not in the set of solved  
goal (respectively node).
The open goal of a node is the first unsolved goal of this node according to 
some preset order. The other unsolved nodes are called pending nodes.
If a node is unsolved, then it contains a unique open goal.
During MCTS exploration, we do not explore pending goals of an unsolved node 
before its open goal is solved which justifies our terminology. In other words, 
our search tree has the following property: if $g$ is a pending node then 
$T(g)$ is empty.

\end{definition}

In Definition~\ref{def:desc}, we define relations between goals 
and nodes in terms of parenthood.

\begin{definition}\label{def:desc}(children, descendant, ancestors)\\
The children of a goal $g$ is a set of nodes defined by:
\[\mathit{Children}(g) = \lbrace A(g,t)\in \mathbb{A}\ |\ t \in T(g) \rbrace\]

@@

\end{definition}

In Definition~\ref{def:prod}, we judge if a tactic is productive from its 
contribution to the search tree.

\begin{definition}\label{def:prof} (Productive tactic)
The application of a tactic on an open goal is called productive if and only if
all the following conditions are satisfied:
\begin{itemize}
\item It does not fail or timeout. To prevent tactics from looping, we
interrupt tactics after a short amount of time (0.05 seconds).
\item It does not loop, i.e its output does not contain any goals that appears 
in the ancestor of the node of the current goal.
\item It is not a parallel step, i.e its output is not a superset
of a list of goals produced from the same goal.
\end{itemize}

@@

\section{Prediction}\label{s:prediction}
The learning-based selection of relevant lemmas significantly improves the
automation for hammers~\cite{BlanchetteGKKU16}. In \tactictoe, we use the
distance-weighted \emph{$k$ nearest-neighbour} classifier~\cite{DudaniS76}
adapted for lemma selection~\cite{ckju-pxtp13}. It allows for hundreds of
predictions per second, while maintaining very high prediction quality~\cite{femalecop}.
We rely on this supervised machine learning algorithm to predict three 
different kinds of objects:
 tactics, theorems and lists of goals. The number of predicted objects $k$ is
 called the \emph{prediction radius}. For a goal $g$, we note
$\mathit{Predictions}^{\mathit{tactic}}_k (g)$ (respectively
$\mathit{Predictions}^{\mathit{theorem}}_k (g)$
and $\mathit{Predictions}^{\mathit{goal\_list}}_k (g)$) the $k$ tactics 
(respectively theorems and lists of goals) selected by our prediction algorithm.

We introduce specificities associated with the prediction of each object.
In particular, we present the dataset from which objects are selected and the
purpose of the selection process. Details on the similarity measure backing the 
predictions is described in Section~\ref{sec:features} and two methods for 
improving the quality of the predictions are presented in 
Section~\ref{sec:ortho} and Section~\ref{sec:synthesis}.

@@

\paragraph{Theorems as Arguments of Tactics}
We first collect a set of theorems for our predictor to select from.
It includes the \holfour theorem database and theorems from the local namespace.
Predicted tactics during proof search and orthogonalization may include
tactics where arguments (lists of theorems) have been erased (see Section
\ref{sec:synthesis}).
We instantiate those tactics by theorems that are syntactically the closest to
the goal as they are more likely to help.
The same algorithm selects suitable premises for ATPs integrated with
\tactictoe (see Section~\ref{sec:atp}).

\paragraph{Lists of Goals as Output of Tactics}
A dataset of tactic outputs is compiled during orthogonalization.
This set is separated into positive and negative examples (see
Section~\ref{sec:evaluation}).
List of goals that are a subset of the result of a recorded tactic on a
targeted goal are considered positive. During proof search, given a list of
goals $l$ created by a tactic, we select a set of lists of goals that are most
similar to $l$. The ratio of positive examples in the selection gives us a
prior evaluation for MCTS.

\section{Proof Search}\label{sec:proofsearch}

Despite the best efforts of the prediction algorithms, a selected tactic may
not solve the current goal, proceed in the wrong direction
or even loop. For that reason, predictions need to be accompanied by a
proof search mechanism that allows for backtrac king and
can choose which proof tree to extend next and in which direction.
Our search algorithm is a Monte Carlo tree search~\cite{} algorithm that relies
on a prior evaluation function and a prior policy function. Those priors are
learned through direct supervised learning for the policy and via the data
accumulated during orthogonalization for the evaluation. Therefore, this MCTS
algorithm does not need to rely on roll outs for evaluation. The
policy and evaluation are estimated by the simple supervised k-NN algorithm
here instead of neural networks trained through reinforcement learning in
AlphaGo Zero~\cite{silver2017mastering}.
A characteristic of the MCTS algorithm is that it offers a good trade-off
between exploitation and exploration. This means that the algorithm searches
deeper more promises branches and leaves enough time for the exploration
of less likely alternatives.

We first describe the proof search algorithm and explain later how to compute 
the prior policy \texttt{PriorPolicy} and prior evaluation 
\texttt{PriorEvaluation} functions.

\subsection{Proof exploration}

The proof search starts by creating an initial search tree. The search tree 
then evolves by the repetitive applications of MCTS steps. A step in the main 
loop of MCTS consists of three parts: node selection, node extension, and 
backpropagation. And the decision to terminate the algorithm is taken during
resolution.

\paragraph{Initialization}
The input of the algorithm is a goal $g$ (also called conjecture) that we 
want to prove.
Therefore the search tree starts with only one node the list of goals
$[g]$ and an ordered list of predicted tactics for the goal $g$.
%Then, the MCTS algorithm alternates node selection steps and node extension
%steps which grow the search tree.

%A similar trade-off between exploration and exploitation exists in algorithm
%like Monte Carlo. However, Monte Carlo tree search metrics are based on many
%simulations and do not reflect our exploration strategy that only
\paragraph{Node selection}
Through node extension steps the search tree grows and the number of paths to
explore increases. To decide which node to extend next, a evolving value for
each node is given in Definition~\ref{def:value}. The algorithm that performs
node selection starts from the root of the search tree. From the current node,
the function \texttt{CompareChildren} choses among the children of its open 
goal the one with the highest value.

If the highest children value is higher than the widening policy, then the 
selected child
becomes the current node, otherwise the final result of node selection is the
current node. The node selection algorithm can be written as the following
pseudo-code:

\begin{definition}\label{def:value}(Value)
The value of the $i^th$ child $a_i$ of the open goal of a parent node $p$ 
is determined by:
\[\mathit{Value}(a_i) = \mathit{CurEvaluation}(a_i) + c_{exploration} * 
\mathit{Exploration}(a_i)\]

The number $\mathit{Failure}(a_i)$ is the amount of failures that occurred 
during node extension from descendants of $a_i$ .
The current evaluation $CurEvaluation(a_i)$ is the average evaluation of
all descendants of $a_i$ including node extension failures.

The exploration term is determined by the prior policy and the current policy.
The current policy is calculate from the number of times a node $x$ has been 
traversed during node selection, noted $\mathit{Visit}(x)$.

\[\mathit{Exploration}(a_i) = 
\frac{\mathit{PriorPolicy}(a_i)}{\mathit{CurPolicy}(a_i)}\]

\paragraph{Node extension}
Let $a$ be the result of node selection.
If $a$ is a solved node or the descendant of a solved node, then extending $a$ 
is useless and the algorithm reports a failure for this MCTS step. 
%In our 
%experiments, filtering such nodes before node selection is detrimental.
If $a$ is not $solved$, it applies the best untested tactic $t$ on the open 
goal $g$ of this
node according to the prediction order for $g$. If no such tactic exists or if
$t$ is not productive, a failure is returned.
If node extension succeeds, a new node containing $t(g)$ is added to the 
children of $a$.

\paragraph{Backpropagation}
During backpropagation we update the statisitics of all the nodes traversed or
created during this MCTS step:
\begin{itemize}
\item Their number of visit is incremented.
\item If node extension failed, their failure count is incremented.
\item If node extension succeeded, they inherit the evaluation of the created
child.
\end{itemize}

\paragraph{Resolution}
The search ends when the algorithm reaches one of these 3 conditions:
\begin{itemize}
\item It finds a proof, i.e. the root node is solved. In this case,
the search returns a minimized and
prettified proof script (see Section~\ref{sec:proofdisplay}).
\item It saturates, i.e. there is no tactics to be applied to any open goal.
This happens less than 10 times in the full-scale experiment.
\item It exceeds a given time limit set by default to 60 seconds.
\end{itemize}

\subsection{Supervised learning guidance}
Our MCTS algorithm relies on the guidance of two supervised learning methods,
the prior policy decides the order in which tactic should be tried and the 
prior evaluation estimates if a branch is fruitful by analysing tactic 
outputs. Both priors influence the rates at which branches of the search tree 
are explored.

\paragraph{Prior policy}\label{sec:policy}
Co-distances between goals produced by the tactic predictor are hard
to translate into a prior policy, so we rely solely on the induced order of
tactics to estimate the probability with which each tactic should be tried.
Let $\mathfrak{T}$ be a search tree after a number of MCTS steps.
Let $p$ be a selected node in $\mathfrak{T}$ and $g$ its open goal. We order 
the list of $n$ productive tactics already
applied to $g$ by their similarity score so that $T(g)= \lbrace
t_0,\ldots,t_{n-1} \rbrace$.
Let $c_{policy}$ be a constant heuristic that estimates how likely a tactic
$t_i$ is to prove the goal.
We can now calculate of a prior policy of a child $a_i$ produced by the tactic
$t_i$ by:
\[PriorPolicy(a_i) = (1 - c_{policy})^{i} * c_{policy}\]

In order to include the possibly of trying more tactics on $g$ we define the
widening policy on the parent $p$ for its open goal $g$ to be:

\paragraph{Prior evaluation}\label{sec:evaluation}

We now concentrate on the definition a reasonable evaluation function for the
output of tactics. Theoretically, an interesting list of goals is one
that has short proofs for each of its goals, and we could estimate their 
lengths from previous proof searches. However, this
approach generates too much data and slows down the proof search. So, we only 
collect output of tactics tested during orthogonalization. We
declare a list of goals $l$ to be positive if it is has been produced by the
winner of an orthogonalization competition. 
The set of positive examples in 
$\mathit{Predictions}^{\mathit{goal\_list}}_k(l)$ is noted 
$\mathit{Positive}^{\mathit{goal\_list}}_k(l)$.
We chose $k=10$ as a default evaluation radius in our experiments.
And we evaluate a node $a$ through the list of goals $G(a)$ it
contains using the prior evaluation function:
\begin{align*}
\mathit{PriorEvaluation}_k (a) &= 
  \frac{card\ \mathit{Positive}^{\mathit{goal\_list}}_k(G(a)}{k}\\
\end{align*}

\subsection{ATP Integration}~\label{sec:atp}
General-purpose proof automation mechanisms which combine proof translation to
ATPs with machine learning (``hammers'') have become quite successful in
enhancing the automation level in proof assistants~\cite{hammers4qed}.
As external automated reasoning techniques sometimes outperform the combined
power of tactics, we would like to combine the \tactictoe search with
general purpose provers such as the ones found in \holyhammer for
\holfour~\cite{tgck-cpp15}.

Such a prover already exists in \holfour and is called \metis. It is already
recognized by the tactic selection mechanism in \tactictoe and can have its
premises predicted dynamically when in an abstracted form. Nevertheless,
we think that the performance of \tactictoe can be boosted by giving the ATP
\metis a special status. This arrangement consists of always predicting
premises for \metis, giving it a slightly higher timeout and
trying it first on each open goal. These modifications only induce an
linear overhead on the number nodes as a general purpose prover does not create
goal obligations.

The integration of an external prover is also possible. And
\eprover~\cite{eprover} is the one we experimented with.
To this end, we created a tactic which expects a number of premises as an 
argument
and calls \eprover on the translated first order problem. We then extract the
premises that were required in the \eprover
proof and reconstruct it inside \holfour with \metis. Giving a special status
to external
provers is essential as external ATP calls do not appear in human
proof scripts. For external ATPs, we use an even higher timeout (5 seconds) and 
a larger number of predicted premises (128 for \eprover) and also try them 
first on open 
goals. Since calls to ATPs are
computationally expensive, they are run in parallel and asynchronously. This
avoids slowing down \tactictoe's search loop. The number of asynchronous calls
to external provers that can be executed at the same time is limited by the
number of cores available to the user.

\section{Feature Extraction}\label{sec:features}
We explain the construction of the similarity measure by which relevant
objects are predicted

We predicts tactics through their associated goals. So, the features 
for each kind of object can be extracted from (different representations of)
mathematical formulas. We start by describing features for \holfour terms. From 
there, we extend the extraction mechanism
to goals, theorems and lists of goals. Duplicated features are always removed
so that each object has an associated set of features.

Here are the different kind of feature we extract from terms:
\begin{itemize}
\item names of constants, including the logical operators,
\item type constructors present in the types of constants and variables,
\item subterms with all variables replaced by a single place holder $V$,
\item names of the variables.
\end{itemize} 
Goals (respectively theorems) are represented by a couple of a list of 
terms (assumptions) and a term (conclusion). So, we distinguish between
features of the assumptions and features of the conclusion by adding a
different tag to elements of each set. The union of all those 
features are considered the feature of a goal (respectively theorem). From 
that, we can construct features for a list of goals by
from the union of the features of each goal in the list.

\subsection{Similarity}\label{sec:predictions}
We estimate the similarity (or co-distance) between two objects $o_1$ and $o_2$
through their respective feature sets $f_1$ and $f_2$.
The estimation is based on the frequency of the features in the intersection of
$f_1$ and $f_2$. A good rule of thumb is that the rarer the shared features
are, the more similar two goals should be. The relative rarity of each feature
can be estimated by calculating its TF-IDF weight~\cite{Jones72astatistical}.
We additionally modify the weight by raising them to the sixth power giving
even more credence
to rare features. This modification was found experimentally optimal in 
\cite{ckju-pxtp13}.

The first co-distance $sim_1$ sums the weight of each shared
features to compute the total score.
In a second co-distance $sim_2$, we additionally take
the total number of features into account, to reduce the seemingly unfair 
advantage of big
feature sets in the first scoring function.
\[sim_1 (o_1, o_2) = {\sum\nolimits_{\,f \in f_0 \cap
<f_1}{\text{tfidf}(f)^{6}}}\]
\[sim_2 (o_1, o_2) = \frac{{\sum\nolimits_{\,f \in f_0 \cap
f_1}{\text{tfidf}(f)^{6}}}}
{ln (e + \mathit{card} f_0 + \mathit{card} f_1)}\]

In our setting, making predictions for an object $o$ consists of sorting a set
of objects by their similarity to $o$. We use $sim_1$ for tactics and theorems
and $sim_2$ for lists of goals. The reason why $sim_1$ is used for
predicting tactics is that tactics effective on a large goal $g$ are often also
suitable for a goal made of sub-formulas of $g$. The same monotonicity
heuristic can justify the use of $sim_1$ for theorems. Indeed, in practice a
large theorem is often a conjunction of theorems from the same domain.
And if a conjunction contains a formula related to a problem, then the other
conjuncts from the same domain may also help to solve it.
